---
title: "Predictor Variables Calculations 4"
date: "2024-04-20"
output:
  html_document:
    code_folding: hide
format:
  html:
    toc: true
    number-sections: true
    colorlinks: true
---

# Clean environment

```{r}
source("../src/A_config.R")  # configurations / data paths / variables / plotting settings / libraries
source("../src/functions.R") # processing functions
```


# Libraries

```{r}

pkg <- c("terra", "ggplot2", "ggfortify")
install_and_load(pkg)


library("rasterSp")
library("climateNiche")

```



```{r}
#| label: get-dependencies-rasterSp-climateNiche
#| eval: false


# Install remotes if not previously installed
if(!"remotes" %in% installed.packages()[,"Package"]) install.packages("remotes")

# Install rasterSp from Github if not previously installed
if(!"rasterSp" %in% installed.packages()[,"Package"]) remotes::install_github("RS-eco/rasterSp", build_vignettes = T)

#Automatically install required packages, which are not yet installed
packages <- c("rgbif", "sp", "raster", "dplyr", "sf", "ggplot2", "rebird", "remotes", "outliers")
new.packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages); rm(new.packages)

# Load packages
l <- sapply(packages, require, character.only = TRUE, quietly=TRUE); rm(packages, l)


# Instal missing packages from GitHub
packages_github <- c("rasterSp", "climateNiche", "ggmap2", "rISIMIP")
new.packages <- packages_github[!(packages_github %in% installed.packages()[,"Package"])]
if(length(new.packages)) remotes::install_github(paste0("RS-eco/", new.packages), build_vignettes=T)
rm(new.packages)

# Load additional packages
l <- sapply(packages_github, require, character.only = TRUE, quietly=TRUE); rm(packages_github, l)
```


# Data

```{r}
# additional path variables for global climate and range data

source_predictors <- paste0(home_path, "Dokumenty/PhD_Projects/StaticPredictors/Data/")
CHELSA_path <- paste0(source_predictors, "Chelsa_Bio_PET/")


# List all the .tif files in the directory
tif_files <- list.files(path=CHELSA_path, pattern = ".tif$")

# The input file geodatabase from BirdLife International 2018
fgdb <-paste0(home_path, "Datasets/Raw/Distributions/Birds_of_the_world/BOTW.gdb")

```

## Species Names

```{r}
# Taxonomic data from BirdLife international
Tax <- st_read(fgdb, layer = "Taxonomic_checklist") %>% 
  dplyr::select(Order_, FamilyName, ScientificName, RL_Category, Synonyms) %>% 
  dplyr::rename("Order" = "Order_") %>%
  mutate(Order = stringr::str_to_sentence(Order))
Tax[Tax == "<NA>"] = NA

# Atlas species names
spData <- read.csv("../../Data/csv/SpeciesNamesData.csv") %>% pull(verbatim_name) 

```

### Create Taxon Lookup table with synonyms from BOTW Taxonomic Info

```{r}
# Reduce Taxonomic data from BirdLife 2018 to Species Names columns
Tax_red <- Tax %>%
    distinct(ScientificName, Synonyms) %>%
    dplyr::rename("AtlasSp" = "ScientificName")
Tax_red$TaxID <- seq(1:nrow(Tax_red))
head(Tax_red)

# Quick Matches
sp_direct_match <- intersect(spData, Tax_red$AtlasSp) # 733
sp_syn_match <- intersect(spData, Tax_red$Synonyms) # 72

# Create Dataframe that will be filled for taxonomic matching
Atlas_sp <- data.frame(
    AtlasSp = spData,
    AtlasID = seq(1:length(spData)),
    match_type = NA,
    match = NA
) %>%
    group_by(AtlasSp) %>%
    mutate(
        match_type = case_when(
            AtlasSp %in% sp_direct_match ~ "direct",
            AtlasSp %in% sp_syn_match ~ "synonym",
            .default = NA
        )
    )

# =================================== Step 1: Direct Matches  ===================================== #

Atlas_merged <- left_join(Atlas_sp, Tax_red %>% filter(AtlasSp %in% sp_direct_match)) %>%
    group_by(AtlasSp) %>%
    mutate(
        match = case_when(is.na(TaxID) ~ 0, !is.na(TaxID) ~ 1),
        AtlasSp_new = case_when(match == 1 ~ AtlasSp, match == 0 ~ ""),
        ScientificName = case_when(match == 0 ~ NA, match == 1 ~ AtlasSp)
    )

Atlas_merged %>% filter(AtlasSp_new == "") %>% ungroup() %>% distinct(match_type)

Atlas_merged %>%
    filter(match == 1) %>%
    n_distinct() # 733 species with matching taxonomy. We can delete them from the matching process now

# Checks:
spData_unmmatched <- Atlas_merged %>%
    filter(match == 0) %>%
    pull(AtlasSp) # 108 species still missing
sp_matched <- Atlas_merged %>%
    filter(match == 1) %>%
    pull(AtlasSp)

syn_match <- intersect(spData_unmmatched, sp_syn_match) # of those 108, 67 species can be matched directly from the synonyms species

intersect(sp_syn_match, sp_matched) # 5 Species were matched but are also found in the synonyms column.

Atlas_merged %>% filter(AtlasSp %in% intersect(sp_syn_match, sp_matched))


# =================================== Step 2: Direct Synonyms =========================================== #
Atlas_synonyms <- Atlas_merged %>%
    dplyr::select(AtlasID, AtlasSp, match_type, match) %>%
    filter(AtlasSp %in% syn_match)
Atlas_synonyms$Synonyms <- Atlas_synonyms$AtlasSp

Atlas_synonyms_merged <- left_join(Atlas_synonyms, Tax_red %>% filter(Synonyms %in% syn_match) %>% rename("ScientificName" = "AtlasSp")) %>%
    mutate(
        AtlasSp_new = ScientificName,
        match = 1
    )
head(Atlas_synonyms_merged)
head(Atlas_merged)
colSums(is.na(Atlas_synonyms_merged))
Atlas_synonyms_merged %>% filter(ScientificName == "")

Atlas_merged2 <- rbind(Atlas_merged %>% filter(!(AtlasSp %in% syn_match)), Atlas_synonyms_merged)

# Delete species from vector that have been matched already:
spData_unmmatched2 <- Atlas_merged2 %>%
    filter(is.na(TaxID)) %>%
    pull(AtlasSp)
Atlas_merged3 <- Atlas_merged2 %>%
    group_by(AtlasSp) %>%
    mutate(
        match_type =
            case_when(
                (AtlasSp %in% spData_unmmatched2) ~ "manually",
                .default = match_type
            )
    )

## Note: There is a script that can do a partial match on the concatenated strings in the original Synonyms column. Not needed here, but just in case.

# =================================== Step 3: Manual Matching  =========================================== #
# Lets continue with manual matching of names: (Note: Updated names were retrieved from Wikipedia on 21.04.2024)
unmatched <- Atlas_merged3 %>% filter(match_type == "manually")
unmatched$AtlasSp_new <- gsub("Sylvia", "Curruca", unmatched$AtlasSp)
unmatched$AtlasSp_new <- gsub("Carduelis cannabina", "Linaria cannabina", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Carduelis chloris", "Chloris chloris", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Linaria chloris", "Chloris chloris", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Locustella pryeri", "Helopsaltes pryeri", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Cyanecula svecica", "Luscinia svecica", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Passer rutilans", "Passer cinnamomeus", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Luscinia akahige", "Larvivora akahige", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Gallirallus okinawae", "Hypotaenidia okinawae", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Luscinia komadori", "Larvivora komadori", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Hirundo daurica", "Cecropis daurica", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Poecile varius", "Sittiparus varius", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Sterna albifrons", "Sternula albifrons", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Egretta intermedia", "Ardea intermedia", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Ammodramus maritimus", "Ammospiza maritima", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Falcipennis canadensis", "Canachites canadensis", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Regulus calendula", "Corthylio calendula", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Oporornis formosus", "Geothlypis formosa", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Bonasa bonasia", "Tetrastes bonasia", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Regulus ignicapillus", "Regulus ignicapilla", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Parus caeruleus", "Cyanistes caeruleus", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Miliaria calandra", "Emberiza calandra", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Delichon urbica", "Delichon urbicum", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Oporornis philadelphia", "Geothlypis philadelphia", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Seiurus noveboracensis", "Parkesia noveboracensis", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Coccothraustes vespertinus", "Hesperiphona vespertina", unmatched$AtlasSp_new)
unmatched$AtlasSp_new <- gsub("Hydropogne caspia", "Hydroprogne caspia", unmatched$AtlasSp_new)

## Back to Step 1: Direct match:
unmatched$ScientificName <- unmatched$AtlasSp_new
unmatched$TaxID <- NULL
unmatched$Synonyms <- NULL

setdiff(unmatched$ScientificName, Tax_red$AtlasSp) # Matched all but 2 species which are new. Great !
unmatched_merged <- left_join(unmatched, Tax_red %>% rename("ScientificName" = "AtlasSp"))
unmatched_merged$match <- 1

## Merge to final table: 
Atlas_merged4 <- rbind(Atlas_merged3 %>% filter(match_type != "manually"), unmatched_merged) %>%
  group_by(AtlasSp) %>%
  mutate()
 Atlas_merged4 %>% filter(match_type == "manually")

## Data Checks: 
colSums(is.na(Atlas_merged4))
intersect(spData, Atlas_merged4$AtlasSp)
unique(Atlas_merged4$ScientificName)
intersect(Atlas_merged4$ScientificName, Tax_red$AtlasSp)
dups <- Atlas_merged4$ScientificName[duplicated(Atlas_merged4$ScientificName)] # 30 duplicates
name_vector <- Atlas_merged4 %>% pull(ScientificName)
Tax_LookUp <- Atlas_merged4 %>% distinct(AtlasSp, ScientificName)
unique(Tax_LookUp$ScientificName)

## Save Output:
Atlas_merged4 %>% 
  filter(ScientificName %in% dups) %>% 
  write.csv("../../Documentation/A_3d_SynonymsAtlasData.csv")

Atlas_merged4 %>%
  write.csv("../../Data/csv/A_3d_full_BirdLife2018_TaxonomicLookup.csv")
  
```

### Get RL Status from BirdLife (tbh this is unneccessary)
```{r}
RedListCat <- Tax %>% 
  filter(ScientificName %in% Tax_LookUp$ScientificName) %>% 
  dplyr::select(ScientificName, RL_Category) %>% 
  left_join(Tax_LookUp) %>% 
  dplyr::select(-ScientificName)

unique(RedListCat$AtlasSp) %>% length()
colSums(is.na(RedListCat))
saveRDS(RedListCat, paste0(out_path, "Predictors/A_3d_RedListCatBL.rds"))


```

### Rasterize BirdLife rangemaps
```{r}
#| label: filter-BirdLife-SQL
#| eval: false

# Create a SQL query to filter the sci_name column and apply additional filters

name_vector_sql <- paste0("'", paste(name_vector, collapse = "','"), "'")

sql_query <- paste0(
  "SELECT sci_name, presence, origin, seasonal, Shape_Length, Shape_Area, Shape ",
  "FROM All_Species ",
  "WHERE sci_name IN (", name_vector_sql, ") ",
  "AND seasonal IN (1, 2) ", # Resident & Breeding Season (not non-breeding or passage)
  "AND origin IN (1, 2) ",   # Native or reintroduced (not introduced or vagrant)
  "AND presence IN (1, 2, 3)" # Extant, probably extant, possibly extant (not possibly extinct or extinct)
)

tictoc::tic()
# Read the filtered data from the geodatabase
All_sp_2 <- suppressMessages(st_read(fgdb, query = sql_query))
tictoc::toc() # takes about 30 minutes

# Save Reduced Version as Shapefile (takes a while)
st_write(All_sp_2, "../../Data/shp/A_3d_BirdLife_810sp.shp", append=F)
```

```{r}
#| label: Rasterize-BirdLife-ranges
#| eval: false


tictoc::tic()
r_birds <- rasterizeRange(dsn= "../../Data/shp/A_3d_BirdLife_810sp.shp",
                          id="sci_nam", touches=TRUE, save = TRUE, resolution = 1,
                          path = "../../Data/species_rasters/")
tictoc::toc() ## Takes about 18 minutes on my local machine

```

### Calculate Global Range Size

::: {.callout-note appearance="minimal"}
## Quote from st_area() function:

if 'sf_use_s2()' is 'FALSE', ellipsoidal distances are computed using st_geod_distance which uses function 'geod_inverse' from GeographicLib (part of PROJ); see Karney, Charles FF, 2013, Algorithms for geodesics, Journal of Geodesy 87(1), 43-55
:::

```{r}
#| eval: false


tictoc::tic()
## Requires BirdLife object from above
# Calculate global range size for each species ## Takes about 11:07 - 
sp <- unique(BirdLife$sci_name)
RangeSizeList <- list()
for (s in seq_along(sp)){
  temp <- BirdLife %>% 
    filter(sci_name == sp[s]) %>% 
    dplyr::select(sci_name)
  
  temp$GlobRangeSize_m2 <- as.numeric(st_area(temp))
  temp$GlobRangeSize_km2 <- temp$GlobRangeSize_m2/1000
  temp2 <- temp %>% st_drop_geometry()
  temp2 <- temp2 %>% 
    group_by(sci_name) %>% 
    mutate(
      GlobRangeSize_m2 = case_when(
        nrow(temp2) > 1 ~ sum(GlobRangeSize_m2),
        nrow(temp) == 1 ~ GlobRangeSize_m2),
      GlobRangeSize_km2 = case_when(
        nrow(temp2) > 1 ~ sum(GlobRangeSize_km2),
        nrow(temp) == 1 ~ GlobRangeSize_km2))
  
  RangeSizeList[[s]] <- temp2[1,]
  
}

RangeSizeBOTW <- plyr::rbind.fill(RangeSizeList) %>% 
  rename("ScientificName" = "sci_name") %>% 
  left_join(Tax_LookUp) %>% 
  select(-ScientificName)

RangeSizeBOTW$GlobRangeSize_km2 <- as.numeric(RangeSizeBOTW$GlobRangeSize_km2 )

head(RangeSizeBOTW)
unique(RangeSizeBOTW$AtlasSp) %>% length()
colSums(is.na(RangeSizeBOTW))


saveRDS(RangeSizeBOTW, paste0(out_path, "Predictors/A_3d_RangeSizeBOTW_df.rds"))

tictoc::toc()
```

```{r}
## Read bird shapefiles back in ===========================================================
sp_v <- stringr::str_split_i(list.files("../../Data/species_rasters/"), 
                           pattern = ".tif", 1)
sp_v <- gsub("_1", "", sp_v) # remove 1 (1 degree resolution)
sp_v <- gsub("_", " ", sp_v)

r_birds <- rast(paste0("../../Data/species_rasters/", 
                       list.files("../../Data/species_rasters/", 
                                  pattern =".tif")))
names(r_birds) <- sp_v
```

## Climate Data

### Aggregate CHELSA to 1° resolution

```{r}
#| eval: false

tictoc::tic()
# read shapefiles:
climate_stack <- rast(paste0(CHELSA_path, tif_files))

## Aggregate to 1 degree cells
target_resolution <- c(120)  # Factor 120 to get from  0.008333° to 1° (110km)
# Aggregate the raster stack to the target resolution
climate_stack_agg <- aggregate(climate_stack, ## This takes ca. 5 minutes on my local machine
                               fact = target_resolution, 
                               fun = mean,
                               na.rm=T)
saveRDS(climate_stack_agg, "../../Data/rds/A_3d_ClimateStack.rds")
tictoc::toc()
```

## Extract Bird Coordinates

```{r}

climate_stack_agg <- readRDS("../../Data/rds/A_3d_ClimateStack.rds")

names(climate_stack_agg) <- c("MAT", 
                              "AnnualPrec", 
                              "PrecWettestMonth", 
                              "PrecDriestMonth", 
                              "PrecSeas", 
                              "MeanDiurnalTempRange", 
                              "Isotherm", 
                              "TempSeas", 
                              "MeanDailyTempWarmestMonth", 
                              "MeanDailyTempColdestMonth", 
                              "AnnualTempRange")

# Extract species coordinates ===========================================================

try({
coords_sp <- list()
for(i in seq_along(sp_v)){
  r_b <- r_birds[[i]]

  # Match resolution, extent, and projection of climate raster to species raster
  r_b_matched <- project(r_b, climate_stack_agg,  method = "average")
  
  r_b2 <- as.data.frame(r_b_matched, xy=T)
  r_b2$species <- sp_v[i]
  r_b3 <- r_b2 %>% rename("Presence" = sp_v[i])
  coords_sp[[i]] <- r_b3
}
})

coords_df <- do.call(rbind,coords_sp)
```

## Climate PCA

```{r}
# Run PCA to build climate space
pca_res <- prcomp(climate_stack_agg, scale. = T)
summary(pca_res)

# extract loadings from PCA for documentation
pca_res$rotation %>% 
  write.csv("../../Documentation/PCA_loadings_climNiche.csv")

```

# Project Birds to climmate space

```{r}
# Loop through species to extract climate values  ======================================
sp_v <- unique(coords_df$species)
clim_vals_list <- list()
pca_all <- list()
for (i in seq_along(sp_v)){
  
  sp <- coords_df %>% filter(species == sp_v[i]) %>% dplyr::select(-species, -Presence)
  clim_vals <- extract(climate_stack_agg, sp, xy=T, df=T) %>% data.frame()
  clim_vals$species = sp_v[i]
  clim_vals_list[[i]] <- clim_vals
  
  # PCA predictions:  ===========================================================
  sp_pca <- predict(pca_res, clim_vals[,2:14])[,1:2] %>% data.frame() %>% mutate(species = sp_v[i])

  pca_all[[i]] <- sp_pca
  
}

clim_vals_all <- do.call(rbind, clim_vals_list) 
colSums(is.na(clim_vals_all))

saveRDS(pca_all, "../../Data/rds/A_3d_pca_all.rds") 

```

## Plot Birds in Climate Space

```{r}
# Plot Climate Space for a couple of species # 
# Step 1: Extract the percentage of variance explained
pca_var <- summary(pca_res)$importance[2, ] * 100

# Step 2: Format the axis labels
x_label <- paste0("PC1 (", round(pca_var[1], 2), "%)")
y_label <- paste0("PC2 (", round(pca_var[2], 2), "%)")

# Step 3: Update the ggplot2 code
library(ggplot2)
library(ggfortify)

ggplot() +
  geom_point(aes(x = PC1, y = PC2), col = "black", data = pca_res) +
  geom_point(aes(x = PC1, y = PC2), col = "#fde725", alpha = 0.6, data = pca_all[[1]]) +
  geom_point(aes(x = PC1, y = PC2), col = "#5ec962", alpha = 0.4, data = pca_all[[2]]) +
  geom_point(aes(x = PC1, y = PC2), col = "#21918c", alpha = 0.3, data = pca_all[[3]]) +
  geom_point(aes(x = PC1, y = PC2), col = "#3b528b", alpha = 0.2, data = pca_all[[4]]) +
  geom_point(aes(x = PC1, y = PC2), col = "#440154", alpha = 0.3, data = pca_all[[5]]) +
  labs(x = x_label, y = y_label) +
  common_theme

ggsave(paste0(out_path, "Figures/BirdsClimateSpace.pdf"))
```

# Calculate Niche Breadth

```{r}
pca_all <-  readRDS("../../Data/rds/A_3d_pca_all.rds") # Predicted values in PCA for species

all_niches <- list()
# Calculate niche breadth
for (i in seq_along(pca_all)){
  niche_breadth2 <- apply(pca_all[[i]][,1:2], 2, sd, na.rm=T) 
  # Create a data frame with species and corresponding variance and standard deviation
  niche_breadth <- data.frame(
    species = unique(pca_all[[i]]$species),
    sd_PC1 = niche_breadth2[1],
    sd_PC2 = niche_breadth2[2])
  all_niches[[i]] <- niche_breadth
}

Niches_df <- plyr::rbind.fill(all_niches)[,1:3] %>% 
rename("ScientificName" = "species") %>% 
mutate(ScientificName = case_when(
  ScientificName == "Mo" ~ "Montifringilla nivalis",
  ScientificName == "Fringilla mo" ~ "Fringilla montifringilla",
  .default = ScientificName)) %>% 
left_join(Tax_LookUp) %>% 
group_by(ScientificName) 

saveRDS(Niches_df, paste0(out_path, "rds/A_3d_Niches_df.rds"))

colSums(is.na(Niches_df))
head(Niches_df)


RangeSizeBOTW <- readRDS(paste0(out_path, "Predictors/A_3d_RangeSizeBOTW_df.rds"))
BirdLife_Predictors <- Niches_df %>% ungroup() %>% select(-ScientificName) %>%
    left_join(RangeSizeBOTW) %>%
    left_join(RedListCat) %>%
    distinct(AtlasSp, .keep_all =T) %>% rename("verbatim_name" = "AtlasSp")

# Data Check: 
BirdLife_Predictors %>% distinct(verbatim_name) %>% nrow()
colSums(is.na(BirdLife_Predictors))


saveRDS(BirdLife_Predictors, paste0(out_path, "Predictors/A_3d_BirdLife2018_Predictors.rds"))

```

## Plot Niche Breadth

```{r}
ggplot(data = Niches_df, aes(x = sd_PC1, y = sd_PC2))+
  geom_point(data = Niches_df, aes(x = sd_PC1, y = sd_PC2, col = AtlasSp), show.legend = F) +
  geom_smooth(method = "lm")+
  common_theme
```
```{r}
save.image(paste0(out_path, "RData/A_3d_ClimateNicheBreadth.RData"))

```

