---
title: "07_WinnersLosers"
format: html
editor: visual
---

this script was cut from 06_Predictions

## Set Color vectors

```{r}
## Color vector for Log Ratio =============================================== ##
my_cols_LR <- c("strong decrease (> halfing)" = "#d7191c",
             "weak decrease (< halfing)" = "#fdae61",
             "stable" = "#e0e0e0",
             "weak increase (< doubling)" = "#a6d96a",
             "strong increase (> doubling)" = "#1a9641")

## Color vector for Jaccard ================================================= ##

my_cols_J <- c("complete turnover (< 0.1)" = "#B35806",
             "strong turnover (0.1 - 0.25)" = "#F1A340",
             "strong intermediate turnover (0.25 - 0.5)" = "#FEE0B6",
             "weak intermediate turnover (0.5 - 0.75)" = "#D8DAEB",
             "weak turnover (> 0.75)" = "#998EC3",
             "stable (> 0.9)" = "#542788")


### Standard Theme ========================================================== ##
common_theme <-   theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), 
    axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 20),
    axis.text = element_text(size = 16),
    axis.text.x = element_text(vjust = 0.9, hjust = 0.9),
    #plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")

```

# \>\> 1. Rank_J_predicted

## 1. Species Ranks (Jaccard Predicted) colored by Log Ratio trend Observed

```{r}

## Czech Republic =========================================================== ##
CZ_p_LR <- Czechia_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_predicted), 
             y = J_Predicted, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Jaccard",       x = "",
    title = "Czech Republic",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), 
    axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## New Yoek State =========================================================== ##
NY_p_LR <- NewYork_df %>%
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_predicted), 
             y = J_Predicted, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Jaccard",       x = "",
    title = "New York State",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), 
    axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## Japan ==================================================================== ##
JP_p_LR <- Japan_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_predicted), 
             y = J_Predicted, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Jaccard",
       x = "",
    title = "Japan",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## Europe =================================================================== ##
EU_p_LR <- Europe_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_predicted), 
             y = J_Predicted, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Jaccard",       x = "",
    title = "Europe",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =2),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")





## Grid them together ======================================================= ##
pdf("../../Results/Figures/3_Predictions/Species_Predictions_coloredByLR.pdf", width = 27, height = 18)
gridExtra::grid.arrange(CZ_p_LR, NY_p_LR, EU_p_LR, JP_p_LR, layout_matrix = rbind(c(1, 2), c(3, 4)))
dev.off()



## Save Figures as PPT ====================================================== ##
# esquisse::ggplot_to_ppt(c("CZ_p_LR", "NY_p_LR", "EU_p_LR", "JP_p_LR"))

```

## 2. Species Ranks (Predicted Jaccard) colored by Jaccard Predicted

```{r}

## Czech Republic =========================================================== ##
CZ_p_J <- Czechia_df %>%
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_predicted), 
             y = J_Predicted, 
             fill = trend_J_Predicted, 
             color = trend_J_Predicted )) +
  geom_col()+
  scale_color_manual(values = my_cols_J) +
  scale_fill_manual(values = my_cols_J) +
  labs(y = "Jaccard",       x = "",
    title = "Czech Republic",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")

## New Yoek State =========================================================== ##

NY_p_J <- NewYork_df %>%
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_predicted), 
             y = J_Predicted, 
             fill = trend_J_Predicted, 
             color = trend_J_Predicted )) +
  geom_col()+
  scale_color_manual(values = my_cols_J) +
  scale_fill_manual(values = my_cols_J) +
  labs(y = "Jaccard",       x = "",
    title = "New York State",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")

## Japan ==================================================================== ##

JP_p_J <- Japan_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_predicted), 
             y = J_Predicted, 
             fill = trend_J_Predicted, 
             color = trend_J_Predicted )) +
  geom_col()+
  scale_color_manual(values = my_cols_J) +
  scale_fill_manual(values = my_cols_J) +
  labs(y = "Jaccard",
       x = "",
    title = "Japan",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")

## Europe =================================================================== ##
EU_p_J <- Europe_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_predicted), 
             y = J_Predicted, 
             fill = trend_J_Predicted, 
             color = trend_J_Predicted )) +
  geom_col()+
  scale_color_manual(values = my_cols_J) +
  scale_fill_manual(values = my_cols_J) +
  labs(y = "Jaccard",       x = "",
    title = "Europe",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =2),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")

## Grid them together ======================================================= ##
pdf("../../Results/Figures/3_Predictions/Species_Predictions_coloredByJ.pdf", width = 27, height = 18)
gridExtra::grid.arrange(CZ_p_J, NY_p_J, EU_p_J, JP_p_J, layout_matrix = rbind(c(1, 2), c(3, 4)))
dev.off()


## Save Figures as PPT ====================================================== ##
# esquisse::ggplot_to_ppt(c("CZ_p_J", "NY_p_J", "EU_p_J", "JP_p_J"))

```

# \>\> 2. Rank_J_observed

## 1. Species Ranks (Jaccard Observed) colored by Jaccard OBSERVED

```{r}

## Czech Republic =========================================================== ##
CZ_p_J_obs <- Czechia_df %>%
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_observed), 
             y = J_Observed, 
             fill = trend_J_Observed, 
             color = trend_J_Observed)) +
  geom_col()+
  scale_color_manual(values = my_cols_J) +
  scale_fill_manual(values = my_cols_J) +
  labs(y = "Jaccard",       x = "",
    title = "Czech Republic",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")

## New Yoek State =========================================================== ##

NY_p_J_obs <- NewYork_df %>%
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_observed), 
             y = J_Observed, 
             fill = trend_J_Observed, 
             color = trend_J_Observed )) +
  geom_col()+
  scale_color_manual(values = my_cols_J) +
  scale_fill_manual(values = my_cols_J) +
  labs(y = "Jaccard",       x = "",
    title = "New York State",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")

## Japan ==================================================================== ##

JP_p_J_obs <- Japan_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_observed), 
             y = J_Observed, 
             fill = trend_J_Observed, 
             color = trend_J_Observed )) +
  geom_col()+
  scale_color_manual(values = my_cols_J) +
  scale_fill_manual(values = my_cols_J) +
  labs(y = "Jaccard",
       x = "",
    title = "Japan",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")

## Europe =================================================================== ##
EU_p_J_obs <- Europe_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_observed), 
             y = J_Observed, 
             fill = trend_J_Observed, 
             color = trend_J_Observed )) +
  geom_col()+
  scale_color_manual(values = my_cols_J) +
  scale_fill_manual(values = my_cols_J) +
  labs(y = "Jaccard",       x = "",
    title = "Europe",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =2),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")

## Grid them together ======================================================= ##
pdf("../../Results/Figures/3_Predictions/Species_Observed_coloredByJ.pdf", width = 27, height = 18)
gridExtra::grid.arrange(CZ_p_J_obs, NY_p_J_obs, EU_p_J_obs, JP_p_J_obs, layout_matrix = rbind(c(1, 2), c(3, 4)))
dev.off()


## Save Figures as PPT ====================================================== ##
# esquisse::ggplot_to_ppt(c("CZ_p_J_obs", "NY_p_J_obs", "EU_p_J_obs", "JP_p_J_obs"))

```

## 2. Species Ranks (J Observed) colored by LR observed

```{r}

## Czech Republic =========================================================== ##
CZ_p_LR_J_observed <- Czechia_df %>%
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_observed), 
             y = J_Observed, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Jaccard",       x = "",
    title = "Czech Republic",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## New Yoek State =========================================================== ##
NY_p_LR_J_observed <- NewYork_df %>%
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_observed), 
             y = J_Observed, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Jaccard",       x = "",
    title = "New York State",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## Japan ==================================================================== ##
JP_p_LR_J_observed <- Japan_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_observed), 
             y = J_Observed, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Jaccard",
       x = "",
    title = "Japan",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## Europe =================================================================== ##
EU_p_LR_J_observed <- Europe_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_J_observed), 
             y = J_Observed, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Jaccard",       x = "",
    title = "Europe",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0.5, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =2),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")





## Grid them together ======================================================= ##
pdf("../../Results/Figures/3_Predictions/Species_Observed_coloredByLR.pdf", width = 27, height = 18)
gridExtra::grid.arrange(CZ_p_LR_J_observed, NY_p_LR_J_observed, EU_p_LR_J_observed, JP_p_LR_J_observed, layout_matrix = rbind(c(1, 2), c(3, 4)))
dev.off()



## Save Figures as PPT ====================================================== ##
# esquisse::ggplot_to_ppt(c("CZ_p_LR_J_observed", "NY_p_LR_J_observed", "EU_p_LR_J_observed", "JP_p_LR_J_observed"))
```

# \>\> 3. Rank_LR_observed

```{r}

## Czech Republic =========================================================== ##
CZ_p_LR_LR_Observed <- Czechia_df %>%
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_LR_observed), 
             y = LR_Observed, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Log Ratio AOO",       x = "",
    title = "Czech Republic",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## New Yoek State =========================================================== ##
NY_p_LR_LR_Observed <- NewYork_df %>%
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_LR_observed), 
             y = LR_Observed, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Log Ratio AOO",       x = "",
    title = "New York State",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## Japan ==================================================================== ##
JP_p_LR_LR_Observed <- Japan_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_LR_observed), 
             y = LR_Observed, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Log Ratio AOO",
       x = "",
    title = "Japan",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## Europe =================================================================== ##
EU_p_LR_LR_Observed <- Europe_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_LR_observed), 
             y = LR_Observed, 
             fill = trend_LR_Observed, 
             color = trend_LR_Observed )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Log Ratio AOO",       x = "",
    title = "Europe",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =2),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")





## Grid them together ======================================================= ##
pdf("../../Results/Figures/3_Predictions/LR_Species_Observed_coloredByLR.pdf", width = 27, height = 18)
gridExtra::grid.arrange(CZ_p_LR_LR_Observed, NY_p_LR_LR_Observed, EU_p_LR_LR_Observed, JP_p_LR_LR_Observed, layout_matrix = rbind(c(1, 2), c(3, 4)))
dev.off()



## Save Figures as PPT ====================================================== ##
# esquisse::ggplot_to_ppt(c("CZ_p_LR_LR_Observed", "NY_p_LR_LR_Observed", "EU_p_LR_LR_Observed", "JP_p_LR_LR_Observed"))
```

# \>\> 4. Rank_LR_predicted

```{r}
## Czech Republic =========================================================== ##
CZ_p_LR_LR_Predicted <- Czechia_df %>%
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_LR_predicted), 
             y = LR_Predicted, 
             fill = trend_LR_Predicted, 
             color = trend_LR_Predicted )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Log Ratio AOO",       x = "",
    title = "Czech Republic",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## New Yoek State =========================================================== ##
NY_p_LR_LR_Predicted <- NewYork_df %>%
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_LR_predicted), 
             y = LR_Predicted, 
             fill = trend_LR_Predicted, 
             color = trend_LR_Predicted )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Log Ratio AOO",       x = "",
    title = "New York State",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## Japan ==================================================================== ##
JP_p_LR_LR_Predicted <- Japan_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_LR_predicted), 
             y = LR_Predicted, 
             fill = trend_LR_Predicted, 
             color = trend_LR_Predicted )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Log Ratio AOO",
       x = "",
    title = "Japan",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =3.8),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")




## Europe =================================================================== ##
EU_p_LR_LR_Predicted <- Europe_df %>% 
  ggplot(aes(x = forcats::fct_reorder(verbatim_name, Rank_LR_predicted), 
             y = LR_Predicted, 
             fill = trend_LR_Predicted, 
             color = trend_LR_Predicted )) +
  geom_col()+
#  scale_color_brewer(palette = "PuOr", direction = 1) +
#  scale_fill_brewer(palette = "PuOr", direction = 1) +
  scale_fill_manual(values = my_cols_LR) +
  scale_color_manual(values = my_cols_LR) +
  labs(y = "Log Ratio AOO",       x = "",
    title = "Europe",
    fill = "Species Trend",
    colour = "Species Trend") +
  geom_hline(yintercept = 0, lwd = 0.1)+
  theme_classic()+
  theme(
    axis.line = element_line(colour = "azure4", linetype = "solid"), axis.ticks = element_line(colour = "gray13"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90, vjust = 0.9, hjust = 0.9, size =2),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, vjust = 2),
    legend.position = "none")





## Grid them together ======================================================= ##
pdf("../../Results/Figures/3_Predictions/LR_Species_Predictions_coloredByLR.pdf", width = 27, height = 18)
gridExtra::grid.arrange(CZ_p_LR_LR_Predicted, NY_p_LR_LR_Predicted, EU_p_LR_LR_Predicted, JP_p_LR_LR_Predicted, layout_matrix = rbind(c(1, 2), c(3, 4)))
dev.off()



## Save Figures as PPT ====================================================== ##
# esquisse::ggplot_to_ppt(c("CZ_p_LR_LR_Predicted", "NY_p_LR_LR_Predicted", "EU_p_LR_LR_Predicted", "JP_p_LR_LR_Predicted"))
```
